Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-10-16 manu-draft.qmd
methods revision from rn comments done

The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series. The EBASE mass balance equation is:

$$
H\frac{dC}{dt} = aPAR - R - k_w\left(C_{Sat} - C \right )
$$ {#eq-ebase}

The parameter $k_w$ is the gas transfer velocity defined as:

@@

For @eq-ebase, the terms on the right side of the equation are gross primary production ($P$, $aPAR$), ecosystem respiration ($R$), and gas exchange ($D$, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth. Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.

@@

@eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange. The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]; fetch and suspended solids have also been considered for estuarine systems [@Borges11]. $k_w$ also increases with temperature because the film of water at the air-sea interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems (e.g., currents, fetch, suspended solids, and wave height), the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14]. However, this parameterization is largely based on open-ocean data and so we allow $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

@@

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $R$ (respiration, $\text{mmol}/\text{m}^2/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT23] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward-in-time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.

@@

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck23]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. Finally, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. All results provided herein were generated using the EBASE R package.

>>>>

EBASE uses a mass balance equation for dissolved oxygen, assuming a well-mixed water column of depth $H$ and no lateral transport processes:

$$
H\frac{dC}{dt} = P - R - D
$$ {#eq-ebase}

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air-water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). The equation has dimensions of moles per unit area of time. $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.

@@

@eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange. The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed. $k_w$ also increases with temperature because the film of water at the air-water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). Changes in model implementation were included in the package as additional changes to BASE. First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The remaining three parameters $a$ , $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters. The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.

@@

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot. Functions in the EBASE R package allow alternative prior distributions from those above.
